The Efficient Handling of Guards \\in the Design of RPython's Tracing JIT

### 2012-09-05 talk/vmil2012/paper.tex
(cfbolz, bivab) address some of the comments by the reviewers

\bigtodo{reviewer c: The use of an interpreter is a typical choice, but it is
not the only choice (e.g., Jikes RVM is compile only) so it might make
sense to introduce this idea earlier.}

Tracing just-in-time (JIT) compilers record and compile commonly executed
linear control flow paths consisting of operations executed by an interpreter.
At points of possible divergence from the traced path operations called guards
are inserted. Furthermore, type guards are inserted to specialize the trace
based on the types observed during tracing. In this paper we describe and
analyze how guards work and explain the concepts used in the intermediate and
low-level representation of the JIT instructions and how these are implemented.
This is done in the context of the RPython language and the PyPy project, which
provides a tracing JIT compiler geared at dynamic language optimization.

@@

RPython can mean one of two things:
\begin{itemize}
 \item the language itself
 \item the translation toolchain used to transform RPython programs to executable units
\end{itemize}
The RPython language is a
statically typed object-oriented high-level subset of Python. The subset is chosen in such a way to make type inference possible\cite{ancona_rpython:_2007}.
The language provides
several features such as automatic memory management
and just-in-time compilation.\todo{reviewer d: I argue that jit compilation is
not a feature of a language, but a feature of a language
\emph{implementation}.}
When writing an interpreter using RPython the
programmer only has to write the interpreter for the language she is
implementing.  The second RPython component, the translation toolchain, is used
to transform the interpreter into a C program.\footnote{
    RPython can also be used to translate programs to CLR and Java
    bytecode~\cite{ancona_rpython:_2007}, but this feature is somewhat
    experimental.
}
During the transformation process
different low level aspects suited for the target environment are automatically
added to the program such as a garbage collector and a tracing JIT compiler.
The process of inserting a tracing JIT is not fully automatic but is guided by
hints from the interpreter author.

@@

\bigtodo{reviewer c: In 2.2 there was a comment that guards can be used to
specialize based on type in dynamic languages. Is this only applicable to
dynamic languages, or could the same thing occur due to simple inheritance?}

@@

After profiling identifies an interesting
path, tracing is started thus recording all operations that are executed on this
path. This includes inlining functional calls.
As in most compilers, tracing JITs use an intermediate representation to
store the recorded operations, typically in SSA
form~\cite{cytron_efficiently_1991}. Since tracing follows actual execution, the
code that is recorded
represents only one possible path through the control flow graph. Points of
divergence from the recorded path are marked with special operations called
\emph{guards}. These operations ensure that assumptions valid during the
tracing phase are still valid when the code has been compiled and is executed.
In the case of dynamic languages, guards are also used to encode type checks
that come from optimistic type specialization by recording the types of
variables seen during tracing\cite{Gal:2009ux}.
After a trace has been recorded it is optimized and then compiled to platform
specific machine code.

@@

\bigtodo{reviewer c: I found the evaluation less detailed than I was hoping
for, which was a shame as I was very interested in learning how well everything
performed after reading the earlier sections!}

@@

All measurements presented in this section do not take garbage collection of resume data and machine code into account. Pieces
of machine code can be globally invalidated or just become cold again. In both
cases the generated machine code and the related data is garbage collected. The
figures show the total amount of operations that are evaluated by the JIT and
the total amount of code and resume data that is generated.

@@

Figure~\ref{fig:benchmarks} summarizes the total number of operations that were
recorded during tracing for each of the benchmarks and what percentage of these
operations are guards. The number of operations was counted on the unoptimized
and optimized traces. The figure also shows the overall optimization rate for
operations, which is between 69.4\% and 83.89\%, of the traced operations and the
optimization rate of guards, which is between 65.8\% and 86.2\% of the
operations. This indicates that the optimizer can remove
most of the guards, but after the optimization pass these still account for
15.2\% to 20.2\% of the operations being compiled and later executed.
The frequency of guard operations makes it important to store the associated
information efficiently and also to make sure that guard checks are executed
quickly.

@@

\bigtodo{reviewer d:
I would also like to see the number of guard that account for 99\% (or even
99.9\%) of all guard failures, in addition to the 50\% number.}

\begin{figure}
    \include{figures/failing_guards_table}
    \caption{Failing guards, guards with more than 200 failures and guards responsible for 50\% of the failures relative to the total number of guards}
    \label{fig:failing_guards}
\end{figure}

@@

The overhead that is incurred by the JIT to manage the resume data,
the backend map as well as the generated machine code is
shown in Figure~\ref{fig:backend_data}. It shows the total memory consumption
of the code and of the data generated by the machine code backend and an
approximation of the size of the resume data structures for the
different benchmarks mentioned above. The machine code taken into account is
composed of the compiled operations, the trampolines generated for the guards
and a set of support functions that are generated when the JIT starts and which
are shared by all compiled traces. The size of the backend map
is the size of the compressed mapping from registers and stack to
IR-level variables and finally the size of the resume data is an
approximation of the size of the compressed high-level resume data as described
in Section~\ref{sec:Resume Data}.\footnote{
The size of the resume data is not measured at runtime, but reconstructed from
log files.}

\bigtodo{ reviewer c: I found it difficult to understand how the size of the
resume structures are approximated, or what you mean in the footnote where you
say it is reconstructed from log files. }

@@

\bigtodo{reviewer c: In 6.1 you claim that the elimination of snapshots is
orthogonal to resume data compression. While they are techniques that can
be used individually or together, I think there may be non-trivial
interactions, in particular eliminating snapshots may alter opportunities
for compression.}

Mike Pall, the author of LuaJIT describes in a post to the lua-users mailing
list different technologies and techniques used in the implementation of
LuaJIT~\cite{Pall:2009}. Pall explains that guards in LuaJIT use a datastucture
called snapshots, similar to RPython's resume data, to store the information
about how to rebuild the state from a guard failure using the information in the
snapshot and the machine execution state. According to Pall~\cite{Pall:2009}
snapshots for guards in LuaJIT are associated with a large memory footprint.
The solution used there is to store sparse snapshots, avoiding the creation
of snapshots for every guard to reduce memory pressure. Snapshots are only
created for guards after updates to the global state, after control flow points
from the original program and for guards that are likely to fail. As an outlook
Pall mentions plans to switch to compressed snapshots to further reduce
redundancy.\footnote{This optimization is now implemented in LuaJIT, at the time of writing it has not been fully documented in the LuaJIT Wiki: \url{http://wiki.luajit.org/Optimizations\#1-D-Snapshot-Compression}}
The approach of not creating snapshots at all for every guard is
orthogonal to the resume data compression presented in this paper and could be
reused within RPython to improve the memory usage further.

@@

For TraceMonkey, a tracing JIT for JavaScript, Gal et. al~\cite{Gal:2009ux}
illustrate how it uses a small off-trace set of instructions that is
executed in case a guard failure to return a structure describing the reason
for the exit along with the information needed to restored the interpreter
state. TraceMonkey uses trace stitching to avoid the overhead of returning to
the trace monitor and calling another trace when taking a side exit. In this
approach it is required to write live values to an activation record before
entering the new trace.

% subsection Guards in Other Tracing JITs (end)

>>>>

Tracing just-in-time (JIT) compilers record and compile commonly executed
linear control flow paths consisting of operations executed by an
interpreter.\footnote{There are also virtual machines that have a tracing JIT
compiler and do not use an interpreter~\cite{xxx}. This paper assumes that the
baseline is provided by an interpreter. Similar design constraints would apply
to a purely compiler-based system.}
At points of possible divergence from the traced path operations called guards
are inserted. Furthermore, type guards are inserted to specialize the trace
based on the types observed during tracing. In this paper we describe and
analyze how guards work and explain the concepts used in the intermediate and
low-level representation of the JIT instructions and how these are implemented.
This is done in the context of the RPython language and the PyPy project, which
provides a tracing JIT compiler geared at dynamic language optimization.

@@

RPython can mean one of two things:
\begin{itemize}
 \item the language itself
 \item the translation toolchain used to transform RPython programs to executable units
\end{itemize}
The RPython language is a
statically typed object-oriented high-level subset of Python. The subset is chosen in such a way to make type inference possible\cite{ancona_rpython:_2007}.
The language tool-set provides
several features such as automatic memory management
and just-in-time compilation.
When writing an interpreter using RPython the
programmer only has to write the interpreter for the language she is
implementing.  The second RPython component, the translation toolchain, is used
to transform the interpreter into a C program.\footnote{
    RPython can also be used to translate programs to CLR and Java
    bytecode~\cite{ancona_rpython:_2007}, but this feature is somewhat
    experimental.
}
During the transformation process
different low level aspects suited for the target environment are automatically
added to the program such as a garbage collector and a tracing JIT compiler.
The process of inserting a tracing JIT is not fully automatic but is guided by
hints from the interpreter author.

@@

@@

After profiling identifies an interesting
path, tracing is started thus recording all operations that are executed on this
path. This includes inlining functional calls.
As in most compilers, tracing JITs use an intermediate representation to
store the recorded operations, typically in SSA
form~\cite{cytron_efficiently_1991}. Since tracing follows actual execution, the
code that is recorded
represents only one possible path through the control flow graph. Points of
divergence from the recorded path are marked with special operations called
\emph{guards}. These operations ensure that assumptions valid during the
tracing phase are still valid when the code has been compiled and is executed.
Guards are also used to encode type checks
that come from optimistic type specialization by recording the types of
variables seen during tracing\cite{Gal:2006, Gal:2009ux}.
After a trace has been recorded it is optimized and then compiled to platform
specific machine code.

@@

@@

All measurements presented in this section do not take garbage collection of resume data and machine code into account. Pieces
of machine code can be globally invalidated or just become cold again. In both
cases the generated machine code and the related data is garbage collected. The
figures show the total amount of operations that are evaluated by the JIT and
the total amount of code and resume data that is generated.
%The measurements and the evaluation focus on trace properties and memory
%consumption, and do not discuss the execution time of the benchmarks. These
%topics were covered in earlier work~\cite{bolz_allocation_2011} and furthermore
%are not influenced that much by the techniques described in this paper.

@@

Figure~\ref{fig:benchmarks} summarizes the total number of operations that were
recorded during tracing for each of the benchmarks and what percentage of these
operations are guards. The static number of operations was counted on the unoptimized
and optimized traces. The figure also shows the overall optimization rate for
operations, which is between 69.4\% and 83.89\%, of the traced operations and the
optimization rate of guards, which is between 65.8\% and 86.2\% of the
operations. This indicates that the optimizer can remove
most of the guards, but after the optimization pass these still account for
15.2\% to 20.2\% of the operations being compiled and later executed.
The frequency of guard operations makes it important to store the associated
information efficiently and also to make sure that guard checks are executed
quickly.

@@

\begin{figure*}
    \include{figures/failing_guards_table}
    \caption{Failing guards, guards with more than 200 failures and guards
    responsible for 50\%, 99\% and 99.9\% of the failures relative to the total number of guards}
    \label{fig:failing_guards}
\end{figure*}

@@

The overhead that is incurred by the JIT to manage the resume data,
the backend map as well as the generated machine code is
shown in Figure~\ref{fig:backend_data}. It shows the total memory consumption
of the code and of the data generated by the machine code backend and an
approximation of the size of the resume data structures for the
different benchmarks mentioned above. The machine code taken into account is
composed of the compiled operations, the trampolines generated for the guards
and a set of support functions that are generated when the JIT starts and which
are shared by all compiled traces. The size of the backend map
is the size of the compressed mapping from registers and stack to
IR-level variables and finally the size of the resume data is
the size of the compressed high-level resume data as described
in Section~\ref{sec:Resume Data}.\footnote{
Due to technical reasons the size of the resume data is hard to measure
directly at runtime. Therefore the size given in the table is reconstructed
from debugging information stored in log files produced by the JIT.}

@@

Mike Pall, the author of LuaJIT describes in a post to the lua-users mailing
list different technologies and techniques used in the implementation of
LuaJIT~\cite{Pall:2009}. Pall explains that guards in LuaJIT use a datastucture
called snapshots, similar to RPython's resume data, to store the information
about how to rebuild the state from a guard failure using the information in the
snapshot and the machine execution state. According to Pall~\cite{Pall:2009}
snapshots for guards in LuaJIT are associated with a large memory footprint.
The solution used there is to store sparse snapshots, avoiding the creation
of snapshots for every guard to reduce memory pressure. Snapshots are only
created for guards after updates to the global state, after control flow points
from the original program and for guards that are likely to fail. As an outlook
Pall mentions plans to switch to compressed snapshots to further reduce
redundancy.\footnote{This optimization is now implemented in LuaJIT, at the time of writing it has not been fully documented in the LuaJIT Wiki: \url{http://wiki.luajit.org/Optimizations\#1-D-Snapshot-Compression}}
It should be possible to combine the approaches of not creating snapshots at all
for every guard and the resume data compression presented in this paper.

@@

For TraceMonkey, a tracing JIT for JavaScript, Gal et. al~\cite{Gal:2009ux}
illustrate how it uses a small off-trace set of instructions that is
executed in case a guard failure to return a structure describing the reason
for the exit along with the information needed to restored the interpreter
state. TraceMonkey uses trace stitching to avoid the overhead of returning to
the trace monitor and calling another trace when taking a side exit. In this
approach it is required to write live values to an activation record before
entering the new trace.
% subsection Guards in Other Tracing JITs (end)
